---
title: "Day 6 Wrangling Tidyverse"
author: "Elke Windschitl"
date: "2022-08-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)

library(tidyverse)
library(janitor)
library(here)
```

## Read in the WB data files

```{r}
wb_indicators <- read_csv(here("data", "wb_indicators.csv"), 
                          na = c("..", ""))
wb_metadata <- read_csv(here("data", "wb_indicators_metadata.csv"))
```

This data set as is has years as headers - use pivot longer to fix. The series name has multiple vaiables in the same column - use pivot wider to fix.

## Tidy up my data

### `pivot_longer()` to get years into single column.

```{r}
wb_indicators_long <- wb_indicators |> 
  pivot_longer(cols = '2001 [YR2001]':'2020 [YR2020]',
               names_to = "year", 
               values_to = "indicator_value")
```

### `separate()` to clean up years column

```{r}
wb_clean <- wb_indicators_long |> 
  separate(col = year,into = c("year", "year_chr"),sep = " ") |> 
  select(-year_chr, -'Country Code', - 'Series Code') |> 
  mutate(year = as.numeric(year)) |> 
  drop_na('Series Name') |> 
  pivot_wider(names_from = 'Series Name', 
              values_from = indicator_value) 

# Widen the variables currently trapped in 'Series Name' to be spread across multiple columns, and get the values that populated the columns from 'indicator_value' (piped above)

# Rename the column headers

names(wb_clean) <- c("country", "year", "access_clean_fuels_pp", "access_electricity_pp", "co2_emissions_kt", "fossil_fuel_cons_pct", "water_stress")

# Make a subset for example
wb_subset <- wb_clean |> 
  filter(country %in% c("Algeria", "Barbados", "Bulgaria", "Chile"))

# Plot
ggplot(data = wb_subset, aes(x = year, y = co2_emissions_kt, group = country))  +
  geom_line(aes(color = country)) +
  facet_wrap(~country) +
  theme_minimal()
```

## Filtering rows

Examples: filter to only keep exact matches

```{r}
ex_1 <- starwars |> 
  filter(height > 180)
```

```{r}
ex_2 <- starwars |> 
  filter(eye_color == "blue")
```

```{r}
ex_3 <- starwars |> 
  filter(homeworld == "Naboo")
```

Examples: filtering to satisfy multiple conditions

```{r}
ex_4 <- starwars |> 
  filter(height > 180 & homeworld == "Tatooine")
```

```{r}
ex_5 <- starwars |> 
  filter(hair_color == "brown" & species == "Human")
```

Examples: filtering OR statement 

```{r}
ex_6 <- starwars |> 
  filter(height > 180 | eye_color == "yellow")
```

```{r}
ex_7 <- starwars |> 
  filter(homeworld == "Endor" | species == "Droid")
```

```{r}
ex_8 <- starwars |> 
  filter(eye_color == "blue" | eye_color == "red" | eye_color == "brown")
#OR
ex_9 <- starwars |> 
  filter(eye_color %in% c("blue", "brown", "red"))



#THIS IS NEVER WHAT YOU WANT TO DO (BELOW)
ex_10 <- starwars |> 
  filter(eye_color == c("blue", "brown", "red"))
#THIS IS TOTALLY WRONG
```

Examples with the negate !

```{r}
ex_12 <- starwars |> 
  filter(species != "Human")
ex_13 <- starwars |> 
  filter(!species %in% c("Human", "Ewok", "Wookiee"))
```

## `select()`: select or exclude columns

Example: select by name

```{r}


ex_14 <- wb_clean |> 
  select(country, year, co2_emissions_kt)
```

```{r}
ex_15 <- wb_clean |> 
  select(year:fossil_fuel_cons_pct)
```

```{r}
ex_16 <- wb_clean |> 
  select(year:fossil_fuel_cons_pct, -access_electricity_pp)
```

```{r}
ex_17 <- wb_clean |> 
  filter(country %in% c("Bolivia", "Chile", "Mexico")) |> 
  select(country:fossil_fuel_cons_pct)
```

```{r}
ex_18 <- wb_clean |> 
  select(year_new = year, 
         emissions = co2_emissions_kt,
         country_name = country)
```

```{r}
ex_19 <- wb_clean |> 
  select(year, country, water_stress) |> 
  filter(country %in% c("Mexico", "Chile", "Argentina"))
```

## Use `rename()` to rename columns in dplyr

```{r}
ex_20 <- wb_clean |> 
  rename(ffcp = fossil_fuel_cons_pct,
         ws = water_stress)
```

```{r}
ex_21 <- wb_clean |> 
  rename(yr = year,
         acfpp = access_clean_fuels_pp)
```

## Use `mutate()` to update an existing column, or add a new one

If you assign a NEW column name, it will add a new column to the end of your data frame. If you assign an existing column name, it will overwrite that column.

```{r}
ex_22 <- wb_clean |> 
  mutate(co2_emissions_t = co2_emissions_kt * 1000) |> 
  relocate(co2_emissions_t, .after = co2_emissions_kt)
```

```{r}
ex_23 <- wb_clean |> 
  mutate(yay = "YAY")
```

```{r}
ex_24 <- wb_clean |> 
  mutate(year = as.character(year))
```

```{r}
ex_25 <- ex_24 |> 
  mutate(year = as.numeric(year))
```

## `dplyr::relocate()`

```{r}
ex_26 <- wb_clean |> 
  relocate(water_stress, .after = year)
```

## Using `group_by() |> summarize`
group_by: Creates groups under the hood based on variables you specify.

summarize: apply your functions to EACH GROUP, and return them in a nice summary table

```{r}
ex_27 <- starwars |> 
  filter(homeworld %in% c("Naboo", "Tatooine")) |> 
  group_by(species) |> 
  summarize(mean_height = mean(height, na.rm = TRUE),
            mean_mass = mean(mass, na.rm = TRUE))
```

```{r}
df <- tribble(
  ~species, ~length, ~number,
  "lobster", 12, 4,
  "lobster", 14, 6,
  "lobster", 15, 2
)

# Creating a new row for every observation of each length bin
df_case <- df |> 
  uncount(number)

# Count counts observations
# Combination of group_by summarize n()

starwars |> 
  count(homeworld, species)

starwars |> 
  group_by(homeworld, species) |> 
  summarize(size = n())
```

Start wb clean
keep obs for Nic, ecu, peru, chile, and for years >2005
add a new column that is co2 emissions convered to megatons

```{r}
ex_28 <- wb_clean |> 
  filter(country %in% c("Nicaragua", "Ecuador", 
                        "Peru", "Chile") & year > 2005) |> 
  select(country, year, co2_emissions_kt) |> 
  mutate(co2_emissions_mt = co2_emissions_kt / 1000)

mt_by_country <- ex_28 |> 
  group_by(country) |> 
  summarise(total_mt = sum(co2_emissions_mt, na.rm = TRUE))
```

